---
---
---
title: "DADA2 ASV table generation from 16S rRNA gene sequences of DNA from Oman samples collected in 2014"
subtitle: "Source file: OM14_dada2_ASV_table_generation.Rmd"
author: "Daniel Nothaft"
date: "`r format(Sys.Date(), '%d %b %Y')`"
output:
  html_document: 
    df_print: paged # omit to disable paged data table output
    css: stylesheet.css # omit if no need for custom stylesheet
    number_sections: yes # change to no for unnumbered sections
    toc: yes # change to no to disable table of contents
    toc_float: true # change to false to keep toc at the top
    toc_depth: 3 # change to specify which headings to include in toc
    code_folding: show # change to hide to hide code by default
editor_options:
  chunk_output_type: inline
---

# Setup
## Load functions and set directories
Set knitting options
```{r knitting-options}
# global knitting options for automatic saving of all plots as .png and .pdf. Also sets cache directory.
knitr::opts_chunk$set(
  dev = c("png", "pdf"), fig.keep = "all",
  dev.args = list(pdf = list(encoding = "WinAnsi", useDingbats = FALSE)),
  fig.path = file.path("fig_output/", paste0(gsub("\\.[Rr]md", "/", knitr::current_input()))),
  cache.path = file.path("cache/", paste0(gsub("\\.[Rr]md", "/", knitr::current_input())))
)
```
Load packages
```{r load-packages, message=FALSE, warning=FALSE}
library(dada2); packageVersion("dada2") # sequence filtering and sample inference with dada2
library(ShortRead); packageVersion("ShortRead") # package for loading, displaying, and handling sequences
library(tidyverse); packageVersion("tidyverse") # handy piping and data operations
library(rjson); packageVersion("rjson") # read json output from figaro
```

```{r source}
# source all relevant scripting files
# paths.R contains paths to local data and programs
source(file.path("scripts", "paths.R"))
```

Set directories
````{r set-project-dir}
# Set up names of sub directories to stay organized
preprocess_path <- file.path(project_path_OM14, "01_preprocess")
    demultiplex_path <- file.path(preprocess_path, "demultiplexed")
        demultiplex_stats_path <- file.path(demultiplex_path, "stats") # statistical summary output from idemp
        demultiplex_unassigned_path <- file.path(demultiplex_path, "unassigned")
        demultiplex_assigned_path <- file.path(demultiplex_path, "assigned")
    figaro_output_path <- file.path(preprocess_path, "Figaro_output")
filter_path <- file.path(project_path_OM14, "02_filter") 
````

## Organize and rename demultiplexing output files

```{r get-demux-output-raw-names}
(demux_output_raw_names <- sort(list.files(demultiplex_path, full.names = FALSE)))
```

Now we read in the names of the demultiplexed and assigned fastq files.
```{r read-fastq-files}
fnFs_raw_names <- sort(list.files(demultiplex_assigned_path, pattern = "R1", full.names = TRUE))
fnRs_raw_names <- sort(list.files(demultiplex_assigned_path, pattern = "R2", full.names = TRUE))
```

Extract sample names for later use
```{r extract-original-sample-names}
# Extract sample names
sample_names_F <- fnFs_raw_names %>% basename() %>% str_remove("Undetermined_S0_L001_R1_001.fastq.gz_") %>% str_remove(".fastq.gz")

sample_names_R <- fnRs_raw_names %>% basename() %>% str_remove("Undetermined_S0_L001_R2_001.fastq.gz_") %>% str_remove(".fastq.gz")

# Check that forward and reverse file names match
if(!identical(sample_names_F, sample_names_R)) stop("Forward and reverse file names do not match.")

# print first few names
sample_names_F %>% head()

# Repeat the above with reverse reads
R1_names_incremented <- vector("character", length(sample_names_F))
for (i in seq_along(R1_names_incremented )) {
  R1_names_incremented[[i]] <- paste0(sample_names_F[[i]], "_S", i, "_L001_R1_001.fastq.gz")
}

# Now, rename files
file.rename(from = fnFs_raw_names, to = file.path(demultiplex_assigned_path, R1_names_incremented))

fnFs <- sort(list.files(demultiplex_assigned_path, pattern = "R1", full.names = TRUE)) # re-assign full names of forward reads, now with Illumina naming convention for use with Figaro.

# Print first few lines of fnFs
fnFs %>% head()

# Repeat the above with reverse reads
R2_names_incremented <- vector("character", length(sample_names_R))
for (i in seq_along(R2_names_incremented )) {
  R2_names_incremented[[i]] <- paste0(sample_names_R[[i]], "_S", i, "_L001_R2_001.fastq.gz")
}

file.rename(from = fnRs_raw_names, to = file.path(demultiplex_assigned_path, R2_names_incremented))

fnRs <- sort(list.files(demultiplex_assigned_path, pattern = "R2", full.names = TRUE))
```

Count and display reads for input
```{r count-input-reads}
# count forward reads
input_read_count_F <- countLines(demultiplex_assigned_path, pattern="_R1_001.fastq")/4

# make forward read counts into tbl with sample name
input_read_count_F_tbl <- input_read_count_F %>% as_tibble(rownames = "file_name") %>% rename(reads_F = value) %>% mutate(sample_id = file_name %>% str_remove(pattern = "_S([[:digit:]]+)_L001_R1_001.fastq.gz"))

# repeat for reverse

input_read_count_R <- countLines(demultiplex_assigned_path, pattern="_R2_001.fastq")/4

input_read_count_R_tbl <- input_read_count_R %>% as_tibble(rownames = "file_name") %>% rename(reads_R = value) %>% mutate(sample_id = file_name %>% str_remove(pattern = "_S([[:digit:]]+)_L001_R2_001.fastq.gz"))


# Join forward and reverse together into one tbl
input_read_count_FR <- input_read_count_F_tbl %>% select(-file_name) %>% left_join(input_read_count_R_tbl %>% select(-file_name), by = "sample_id") %>% select(sample_id, everything())

# print. reads counts are same in forward and reverse, which is what we want.
input_read_count_FR
```
Plot read counts
```{r plot-input-reads}
plot_read_count_input <- input_read_count_FR %>% ggplot(aes(
  x = fct_reorder(sample_id, desc(reads_F)),
  y = reads_F
)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(name = "Forward reads") +
  scale_x_discrete(name = "Sample ID") +
  theme_classic(base_size = 9)+
  theme(
    axis.text.x = element_text(angle = 90, vjust = .5, hjust = 1),
    legend.position = "bottom"
  )

plot_read_count_input
```

# Check primer sequences

The 515 (forward) (Caporaso) and 806R (reverse) (Caporaso) primers ([EarthMicrobiome](https://earthmicrobiome.org/protocols-and-standards/16s/)) were used to amplify this dataset. We record the DNA sequences for those primers.

However, the sequencing strategy used here is not supposed to sequence the primers.
```{r primers-515F–806R, warning=FALSE, message=FALSE}
FWD <- "GTGCCAGCMGCCGCGGTAA" # 515F (Caporaso)
REV <- "GGACTACHVGGGTWTCTAAT" # 806R (Caporaso)

(FWD_length <- str_length(FWD))
(REV_length <- str_length(REV))
```

That said, let's check for any primer sequence matches in the sequences.

```{r Primer-function, warning=FALSE, message=FALSE, tidy=TRUE}
allOrients <- function(primer){ # Create all orientations of the input sequence
  require(Biostrings)
  dna <- DNAString(primer) # The Biostrings works w/ DNAString objects rather than character vectors
  orients <- c(Forward=dna, Complement=complement(dna), Reverse=reverse(dna), RevComp=reverseComplement(dna))
  return(sapply(orients, toString)) # Convert back to character vector
}
FWD.orients <- allOrients(FWD)
REV.orients <- allOrients(REV)
FWD.orients
```

We are now ready to count the number of times the primers appear in the forward and reverse read, while considering all possible primer orientations. Identifying and counting the primers on one set of paired end FASTQ files is sufficient, assuming all the files were created using the same library preparation, so we'll just process the first sample.

```{r primerHits}
# write function to check for primer sequences
primerHits <- function(primer, fn) { # Counts number of reads in which the primer is found
    nhits <- vcountPattern(primer, sread(readFastq(fn)), fixed = FALSE)
    return(sum(nhits > 0))
}
```

```{r count-primers-1}
# print file name
print(paste("file_name: ", basename(fnFs[[1]])))

print(paste("total reads:", countLines(fnFs[[1]])/4))

# check for primers
rbind(FWD.ForwardReads = sapply(FWD.orients, primerHits, fn=fnFs[[1]]),
      FWD.ReverseReads = sapply(FWD.orients, primerHits, fn=fnRs[[1]]),
      REV.ForwardReads = sapply(REV.orients, primerHits, fn=fnFs[[1]]),
      REV.ReverseReads = sapply(REV.orients, primerHits, fn=fnRs[[1]]))
```

```{r count-primers-2}
# print file name
print(paste("file_name: ", basename(fnFs[[2]])))

print(paste("total reads:", countLines(fnFs[[2]])/4))

# check for primers
rbind(FWD.ForwardReads = sapply(FWD.orients, primerHits, fn=fnFs[[2]]),
      FWD.ReverseReads = sapply(FWD.orients, primerHits, fn=fnRs[[2]]),
      REV.ForwardReads = sapply(REV.orients, primerHits, fn=fnFs[[2]]),
      REV.ReverseReads = sapply(REV.orients, primerHits, fn=fnRs[[2]]))
```
There were a some (<300) detections FWD and REV primer sequences in the forward and reverse reads in its both normal and complement orientation. We will check if primer sequences are still found later after we truncate lower quality reads. So now let's inspect read quality profiles.

# Inspect read quality profiles
It's important to get a feel for the quality of the data that we are using. To do this, we will plot the quality of some of the samples.

*From the dada2 tutorial:*
>In gray-scale is a heat map of the frequency of each quality score at each base position. The median quality score at each position is shown by the green line, and the quartiles of the quality score distribution by the orange lines. The red line shows the scaled proportion of reads that extend to at least that position (this is more useful for other sequencing technologies, as Illumina reads are typically all the same lenghth, hence the flat red line).

````{r plot-qual-R1, fig.width = 13, fig.height = 7}
# Forward reads
# If the number of samples is 20 or less, plot them all, otherwise, just plot 20 randomly selected samples
if( length(fnFs) <= 20) {
  plotQualityProfile(fnFs)
} else {
  rand_samples <- sample(size = 20, 1:length(fnFs)) # grab 20 random samples to plot
  plotQualityProfile(fnFs[rand_samples])
}
````

````{r plot-qual-R2, fig.width = 13, fig.height = 7}
# Reverse reads
# If the number of samples is 20 or less, plot them all, otherwise, just plot 20 randomly selected samples
if( length(fnRs) <= 20) {
  plotQualityProfile(fnRs)
} else {
  rand_samples <- sample(size = 20, 1:length(fnRs)) # grab 20 random samples to plot
  plotQualityProfile(fnRs[rand_samples])
}
````

# Filter parameter optimization with Figaro
Create a directory for Figaro output
```{r create-dir-figaro}
# Create directory to hold the output from figaro
if (!dir.exists(figaro_output_path)) dir.create(figaro_output_path)
```
I couldn't get figaro (v1.1.1) to run by system2() in R due to a shell permission denied error, but it worked on the command line.
This chunk generates the exact text I entered in the command line after changing to the figaro directory. I then just read in the figaro output JSON and continue from there.
```{r cd-figaro}
# text to change directory in terminal to figaro 
paste("cd", figaro_path) %>% str_remove("/figaro.py") %>% cat()
```

```{r figaro}
# primers were not sequenced
# sequences are 1X140 bp
# Tried 240, lost nearly all reads at merge step.
# Try again with 260
figaro_amplicon_length <- 260

figaro_min_overlap <- 20

figaro_output_file_name <- "Figaro_params_OM14.json"

# using 1 for primer lengths because even though primers were not sequenced, the current version of figaro does not allow 0 as a primer length.

flags_figaro <- paste("--ampliconLength", figaro_amplicon_length, "--forwardPrimerLength", 1, "--reversePrimerLength", 1, "--outputFileName", figaro_output_file_name, "--inputDirectory", demultiplex_assigned_path, "--outputDirectory", figaro_output_path, "--minimumOverlap", figaro_min_overlap)

paste("python3", "figaro.py", flags_figaro) %>% cat()
```

Read in Figaro JSON output, top 5 suggested trim parameters.
```{r read-figaro-json}
# Give the input file name to the function.
figaro_result <- fromJSON(file = file.path(figaro_output_path, figaro_output_file_name))

figaro_result[1:5] %>% paste(collapse = '\n') %>% cat()
```

```{r save-figaro-best-params}
# Save and print Figaro-suggested parameters for dada2::filterAndTrim
(truncLen_best_figaro <- figaro_result[[1]]$trimPosition)

(maxEE_best_figaro <- figaro_result[[1]]$maxExpectedError)
```

# Filter and trim for quality

## Organize file paths and directories
````{r set-filtering-file-paths}
# set file path for quality-filtered read
filter_path <- file.path(project_path_OM14, "02_filter") 

# Put filtered reads into separate sub-directories for big data workflow

if (!dir.exists(filter_path)) dir.create(filter_path)
    subF_path <- file.path(filter_path, "preprocessed_F") 
    subR_path <- file.path(filter_path, "preprocessed_R") 
if (!dir.exists(subF_path)) dir.create(subF_path)
if (!dir.exists(subR_path)) dir.create(subR_path)
    
# Move demultiplexed and assigned R1 and R2 to separate forward/reverse sub-directories

fnFs.Q <- file.path(subF_path,  basename(list.files(demultiplex_assigned_path , pattern="R1", full.names = TRUE))) 
fnRs.Q <- file.path(subR_path,  basename(list.files(demultiplex_assigned_path , pattern="R2", full.names = TRUE)))

file.rename(from = list.files(demultiplex_assigned_path , pattern="R1", full.names = TRUE), to = fnFs.Q)
file.rename(from = list.files(demultiplex_assigned_path , pattern="R2", full.names = TRUE), to = fnRs.Q)

# File parsing; create file names and make sure that forward and reverse files match
filtpathF <- file.path(subF_path, "filtered") # files go into preprocessed_F/filtered/
filtpathR <- file.path(subR_path, "filtered") # ...
fastqFs <- sort(list.files(subF_path, pattern="fastq.gz"))
fastqRs <- sort(list.files(subR_path, pattern="fastq.gz"))
if(length(fastqFs) != length(fastqRs)) stop("Forward and reverse files do not match.")
````


Before chosing sequence variants, we want to trim reads where their quality scores begin to drop (the `truncLen` and `truncQ` values) and remove any low-quality reads that are left over after we have finished trimming (the `maxEE` value).

## Filter and trim

````{r filterAndTrim}
# No argument pasted to trimLeft because the primers were not sequenced
filt_out <- filterAndTrim(fwd = file.path(subF_path, fastqFs), filt = file.path(filtpathF, fastqFs), rev = file.path(subR_path, fastqRs), filt.rev = file.path(filtpathR, fastqRs), truncLen = truncLen_best_figaro, maxEE = maxEE_best_figaro, truncQ = 2, maxN = 0, rm.phix = TRUE, compress = TRUE, verbose = TRUE, multithread = TRUE)

# look at how many reads were kept
head(filt_out)

# summary of samples in filt_out by percentage
filt_out %>% 
  data.frame() %>% 
  mutate(Samples = rownames(.),
         percent_kept = 100*(reads.out/reads.in)) %>%
  select(Samples, everything()) %>%
  summarise(min_remaining = paste0(round(min(percent_kept), 2), "%"), 
            median_remaining = paste0(round(median(percent_kept), 2), "%"),
            mean_remaining = paste0(round(mean(percent_kept), 2), "%"), 
            max_remaining = paste0(round(max(percent_kept), 2), "%"))
````

## Check if primers were removed by filterAndTrim
Generate sorted file name vector for filtered files
```{r}
fnFs_filt <- sort(list.files(filtpathF, full.names = TRUE))

# print first 6 lines
fnFs_filt %>% head()

fnRs_filt <- sort(list.files(filtpathR, full.names = TRUE))
```

```{r count-primers-filtered-1}
# print file name
print(paste("file_name: ", basename(fnFs_filt[[1]])))

print(paste("total reads:", countLines(fnFs_filt[[1]])/4))

# tally primer hits in file
rbind(FWD.ForwardReads = sapply(FWD.orients, primerHits, fn=fnFs_filt[[1]]),
      FWD.ReverseReads = sapply(FWD.orients, primerHits, fn=fnFs_filt[[1]]),
      REV.ForwardReads = sapply(REV.orients, primerHits, fn=fnFs_filt[[1]]),
      REV.ReverseReads = sapply(REV.orients, primerHits, fn=fnFs_filt[[1]]))
```

```{r count-primers-filtered-2}
# print file name
print(paste("file_name: ", basename(fnFs_filt[[2]])))

print(paste("total reads:", countLines(fnFs_filt[[2]])/4))

# tally primer hits in file
rbind(FWD.ForwardReads = sapply(FWD.orients, primerHits, fn=fnFs_filt[[2]]),
      FWD.ReverseReads = sapply(FWD.orients, primerHits, fn=fnFs_filt[[2]]),
      REV.ForwardReads = sapply(REV.orients, primerHits, fn=fnFs_filt[[2]]),
      REV.ReverseReads = sapply(REV.orients, primerHits, fn=fnFs_filt[[2]]))
```

Primer sequences appear in ~1 in every 10 000 reads. At such low primer counts, additional filtering of primers is probably not needed, and won't be performed here.

## Plot the quality of the filtered and cut fastq files.

````{r plot-qual-R1-filt, fig.width = 13, fig.height = 7}
# Forward reads
# If the number of samples is 20 or less, plot them all, otherwise, just plot 20 randomly selected samples
if( length(fnFs_filt) <= 20) {
  plotQualityProfile(fnFs_filt)
} else {
  rand_samples <- sample(size = 20, 1:length(fnFs_filt)) # grab 20 random samples to plot
  plotQualityProfile(fnFs_filt[rand_samples])
}
````

````{r plot-qual-R2-filt, fig.width = 13, fig.height = 7}
# Forward reads
# If the number of samples is 20 or less, plot them all, otherwise, just plot 20 randomly selected samples
if( length(fnFs_filt) <= 20) {
  plotQualityProfile(fnFs_filt)
} else {
  rand_samples <- sample(size = 20, 1:length(fnFs_filt)) # grab 20 random samples to plot
  plotQualityProfile(fnFs_filt[rand_samples])
}
````

# Infer sequence variants
In this part of the pipeline, dada2 will learn to distinguish error from biological 
differences using a subset of our data as a training set. After it understands the 
error rates, we will reduce the size of the dataset by combining all identical 
sequence reads into "unique sequences". Then, using the dereplicated data and 
error rates, dada2 will infer the sequence variants in our data. Finally, 
we will merge the corresponding forward and reverse reads to create a list of the 
fully denoised sequences and create a sequence table from the result.

## Learn the error rates

````{r learn-error-rates}
set.seed(100) # set seed to ensure that randomized steps are reproducible

# Learn forward error rates
errF <- learnErrors(fnFs_filt, nbases=1e8, multithread=TRUE)

# Learn reverse error rates
errR <- learnErrors(fnRs_filt, nbases=1e8, multithread=TRUE)
````

### Plot Error Rates
The error rates for each possible transition (A→C, A→G, …) are shown. Points are the observed error rates for each consensus quality score. The black line shows the estimated error rates after convergence of the machine-learning algorithm. The red line shows the error rates expected under the nominal definition of the Q-score. Here, we want to see estimated error rates (black line) that are a good fit to the observed rates (points), and that the error rates drop with increased quality as expected. If these criteria aren't observed, then it may be a good idea to increase the ```nbases``` parameter. This allows the machine learning algorithm to train on a larger portion of your data and may help improve the fit.

````{r plotErrors-f}
plotErrors(errF, nominalQ=TRUE)
```

````{r plotErrors-R}
plotErrors(errR, nominalQ=TRUE)
````

## Dereplication, sequence inference, and merging of paired-end reads
Extract sample names and add them to vector of filtered file names
```{r extract-sample-names}
# Extract sample names, assuming filenames have format: SAMPLENAME_XXX.fastq
sample_names_F_filt <- fnFs_filt %>% basename() %>% str_remove(pattern = "_S([[:digit:]]+)_L001_R1_001.fastq.gz")
sample_names_R_filt <- fnRs_filt %>% basename() %>% str_remove(pattern = "_S([[:digit:]]+)_L001_R2_001.fastq.gz")

# Double check
if(!identical(sample_names_F_filt, sample_names_R_filt)) stop("Forward and reverse files do not match.")

names(fnFs_filt) <- sample_names_F_filt
names(fnRs_filt) <- sample_names_R_filt
```

````{r derep-dada}
# make lists to hold the loop output
mergers <- vector("list", length(sample_names_F))
names(mergers) <- sample_names_F
ddF <- vector("list", length(sample_names_F))
names(ddF) <- sample_names_F
ddR <- vector("list", length(sample_names_F))
names(ddR) <- sample_names_F

# For each sample, get a list of merged and denoised sequences
for(sam in sample_names_F) {
    cat("Processing:", sam, "\n")
    # Dereplicate forward reads
    derepF <- derepFastq(fnFs_filt[[sam]])
    # Infer sequences for forward reads
    ddF[[sam]] <- dada(derepF, err=errF, multithread=TRUE)
    # Dereplicate reverse reads
    derepR <- derepFastq(fnRs_filt[[sam]])
    # Infer sequences for reverse reads
    ddR[[sam]] <- dada(derepR, err=errR, multithread=TRUE)
    # Merge reads together
    merger <- mergePairs(ddF[[sam]], derepF, ddR[[sam]], derepR)
    mergers[[sam]] <- merger
}

rm(derepF); rm(derepR)
````

## Construct sequence table

````{r construct-seqtab}
seqtab <- makeSequenceTable(mergers)
```

```{r write-seqtab}
write_rds(seqtab, path = format(Sys.Date(), "data_output/seqtab_OM14_processed_%Y%m%d.rds"), compress = "gz")
```

# Remove Chimeras
Although dada2 has searched for insertion/deletion errors and substitutions, there may still be chimeric
sequences in our dataset (sequences that are derived from forward and reverse sequences from 
two different organisms becoming fused together during PCR and/or sequencing). To identify 
chimeras, we will search for rare sequence variants that can be reconstructed by combining
left-hand and right-hand segments from two more abundant "parent" sequences.

````{r rm-chimeras}
# Remove chimeras
seqtab_nochim <- removeBimeraDenovo(seqtab, method="consensus", multithread=TRUE)
```

````{r percent-non-chimeric}
# Print percentage of our sequences that were not chimeric.
100*sum(seqtab_nochim)/sum(seqtab)
```

```{r write-seqtab-nochim}
write_rds(seqtab_nochim, path = format(Sys.Date(), "data_output/seqtab_nochim_OM14_processed_%Y%m%d.rds"), compress = "gz")
```

# Track reads through pipeline
```{r track}
getN <- function(x) sum(getUniques(x)) # function to grab sequence counts from output objects
# tracking reads by counts
filt_out_track <- filt_out %>%
  data.frame() %>%
  mutate(Sample = rownames(.) %>% str_remove(pattern = "_S([[:digit:]]+)_L001_R1_001.fastq.gz")) %>%
  rename(input = reads.in, filtered = reads.out)
rownames(filt_out_track) <- filt_out_track$Sample

ddF_track <- data.frame(denoisedF = sapply(ddF[sample_names_F_filt], getN)) %>%
  mutate(Sample = row.names(.))
ddR_track <- data.frame(denoisedR = sapply(ddR[sample_names_F_filt], getN)) %>%
  mutate(Sample = row.names(.))
merge_track <- data.frame(merged = sapply(mergers, getN)) %>%
  mutate(Sample = row.names(.))
chim_track <- data.frame(nonchim = rowSums(seqtab_nochim)) %>%
  mutate(Sample = row.names(.))

track <- left_join(filt_out_track, ddF_track, by = "Sample") %>%
  # left_join(ddF_track, by = "Sample") %>%
  left_join(ddR_track, by = "Sample") %>%
  left_join(merge_track, by = "Sample") %>%
  left_join(chim_track, by = "Sample") %>%
  replace(., is.na(.), 0) %>%
  select(Sample, everything())
row.names(track) <- track$Sample
head(track)

# tracking reads by percentage
track_pct <- track %>% 
  data.frame() %>%
  mutate(Sample = rownames(.),
         filtered_pct = ifelse(filtered == 0, 0, 100 * (filtered/input)),
          # cut_pct = ifelse(cut == 0, 0, 100 * (cut/filtered)),
         denoisedF_pct = ifelse(denoisedF == 0, 0, 100 * (denoisedF/filtered)),
         denoisedR_pct = ifelse(denoisedR == 0, 0, 100 * (denoisedR/filtered)),
         merged_pct = ifelse(merged == 0, 0, 100 * merged/((denoisedF + denoisedR)/2)),
         nonchim_pct = ifelse(nonchim == 0, 0, 100 * (nonchim/merged)),
         total_pct = ifelse(nonchim == 0, 0, 100 * nonchim/input)) %>%
  select(Sample, ends_with("_pct"))

# summary stats of tracked reads averaged across samples
track_pct_avg <- track_pct %>% summarize_at(vars(ends_with("_pct")), 
                                            list(avg = mean))
head(track_pct_avg)


track_pct_med <- track_pct %>% summarize_at(vars(ends_with("_pct")), 
                                            list(avg = stats::median))
head(track_pct_avg)

head(track_pct_med)

# Plotting each sample's reads through the pipeline
track_plot <- track %>% 
  pivot_longer(-Sample, names_to = "Step", values_to = "Reads") %>% 
  mutate(Step = factor(Step, 
                       levels = c("input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim"))) %>%
  ggplot(aes(x = Step, y = Reads)) +
  geom_line(aes(group = Sample), alpha = 0.2) +
  geom_point(alpha = 0.5, position = position_jitter(width = 0)) + 
  stat_summary(fun = median, geom = "line", group = 1, color = "steelblue", size = 1, alpha = 0.5) +
  stat_summary(fun = median, geom = "point", group = 1, color = "steelblue", size = 2, alpha = 0.5) +
  stat_summary(fun.data = median_hilow, fun.args = list(conf.int = 0.5), 
               geom = "ribbon", group = 1, fill = "steelblue", alpha = 0.2) +
  geom_label(data = t(track_pct_avg[1:5]) %>% data.frame() %>% 
               rename(Percent = 1) %>%
               mutate(Step = c("filtered", "denoisedF", "denoisedR", "merged", "nonchim"),
                      Percent = paste(round(Percent, 2), "%")),
             aes(label = Percent), y = 1.1 * max(track[,2])) +
  geom_label(data = track_pct_avg[6] %>% data.frame() %>%
               rename(total = 1),
             aes(label = paste("Total\nRemaining:\n", round(track_pct_avg[1,6], 2), "%")), 
             y = mean(track[,6]), x = 6.6) +
  expand_limits(y = 1.1 * max(track[,2]), x = 7.2) +
  theme_classic()

track_plot
```
